A Practical Guide for Generating Reproducible and Programmatic Neuroimaging Visualizations

### 2023-06-20 manuscript.Rmd
rdme sp fc
revised2

---
title: A Practical Guide for Generating Reproducible and Programmatic Neuroimaging Visualizations
authors:
  - name: Sidhant Chopra
    thanks: Corresponding author
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
    email: sidhant.chopra@yale.com
  - name: Loïc Labache
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Elvisha Dhamala
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Edwina R Orchard
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Avram Holmes
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
keywords:
  - Neuroimaging visualization
  - Reproducibility 
  - Programmatic figures
  - Open Science
  - R
  - Python
  - Brain Visualization
abstract: Neuroimaging visualizations form the centerpiece of the interpretation and communication of scientific results, and are also a cornerstone for data quality control. Often, these images and figures are produced by manually changing settings on Graphical User Interfaces (GUIs). There now exist many well-documented code-based brain visualization tools that allow users to programmatically generate publication-ready figures directly within R, Python and MATLAB environments. Here, we provide a rationale for the wide-spread adoption of code-generated brain visualizations by highlighting corresponding advantages in replicability, flexibility, and integration over GUI based tools. We then provide a practical guide outlining the steps required to generate these code-based brain visualizations. We also present a comprehensive table of tools currently available for programmatic brain visualizations and provide didactic examples of visualizations and associated code as a point of reference. Finally, we provide a web-app (https://sidchop.shinyapps.io/braincode/) which can generate simple code-templates as starting points for these visualizations.
output: 
  bookdown::word_document2:
    toc: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
    keep_tex: true
    latex_engine: lualatex
linestretch: 1.5
geometry: "left=1.5cm, right=1.5cm, top=1.5cm, bottom=1.5cm"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{setspace}\onehalfspacing
- \usepackage[left]{lineno}
- \modulolinenumbers[5]
- \linenumbers
---

@@

The visualization of neuroimaging data is one of the primary ways in which we evaluate data quality, interpret results, and communicate findings. These visualizations are commonly produced using graphical user interface-based (GUI) tools where individual images are opened and, within each instance, display settings are manually changed until the desired output is reached. In large part, the choice to use GUI-based software has been driven by a perception of convenience, flexibility, and accessibility. However, there now exist many code-based software packages which are well-documented and often do not require high-level knowledge of programming, making them more accessible to the neuroimaging community. These tools are flexible and allow for the generation of reproducible, high-quality, and publication-ready brain visualizations in only a few lines of code, especially within the R, Python and MATLAB environments. Here, we first discuss the rationale for the widespread adoption of code-generated visualizations by highlighting major advantages in replicability, flexibility, and integration over GUI based tools. We then provide a practical guide outlining the steps required to make code-based brain visualizations and provide a web-app (<https://sidchop.shinyapps.io/braincode/>; Figure 4) that can generate simple code-templates as starting points for these visualizations. We also present a comprehensive table of tools currently available for programmatic brain visualizations (Table 1) and provide didactic examples of visualizations and associated code as a point of reference (Figure 2-3). The focus of this guide is on human brain magnetic resonance imaging (MRI) data, but many of the principles discussed and tools provided will equally apply to visualizing data from other organs and imaging modalities such as EEG, MEG, PET and CT.

@@

Notably, some GUI-based tools have historically offered command-line access to generate replicable visualizations (e.g., FreeView, FSLeyes, surfice), making their use potentially equally replicable to purely code-based tools. Use of these specialized command-line interfaces can provide a useful middle ground of those who have little experience with coding environments. Nonetheless, these interfaces often still have a learning curve, but can lack other advantages, such as iteration, provided by the preferred programming environment (see Sections \@ref(flexibility) and \@ref(integrative-and-interactive-reporting)). Likewise, other GUI-based tools offer replicability in the form of automatically generated batch scripts (text files containing lines of specialized commands that can be re-executed) or in-built terminals, which are often idiosyncratic and may lack documentation to make them easily usable or replicable by those not familiar with the specific software.

## Flexibility

Being able to exactly replicate a figures via code has marked advantages beyond open science practices. In particular, the ability to reprogram inputs (such as statistical maps) and settings (such as color schemes, thresholds, and visual orientations) can streamline the entire scientific workflow. Changing inputs and settings via code allows for the easy production of multiple figures, such as those resulting from multiple analyses which require similar visualizations. A simple for-loop or copying and pasting the code with altered input and/or settings-of-interest can be a powerful method for exploring visualization options or rapidly creating multi-panel figures. Likewise, an arduous request from a reviewer or collaborator to alter the image processing or analysis becomes less of a burden when the associated figures can be re-generated with a few lines of code, as opposed to re-pasting and re-illustrating them manually. Having a code-base with modifiable inputs can mean that the generation of visualizations requires less time, energy and effort than image and instance specific GUI-based generation. This also makes it easier to generate consistent figures across subsequent projects. Keep in mind that the gains of writing code for figures are cumulative, and in addition to improving programming skills, one can build a code-base for figure generation that can continue to be reused and shared throughout a scientific career.

Precise controls via code over visualization settings, such as color schemes, legend placement and camera angles, can provide much greater flexibility over visualizations. Nonetheless, part of the appeal of GUI-based tools is that the presets for such settings can provide a useful starting point and reduce the decision burden on novice users. However, similar presets are often available in the form of default settings across most code-based packages, negating the need for the user to manually enter each and every choice required for creating an image. Most code-based tools also come with documentation, with R-packages on the *CRAN* or *Neuroconductor* (Muschelli et al., 2019) repositories strictly requiring detailed guidance. Recent packages have started to include detailed beginner-friendly documentation in GitHub repositories, or even entire papers (e.g., Pham, Muschelli, & Mejia, 2022; Mowinckel & Vidal-Piñeiro, 2020; Huntenburg et at., 2017; Schäfer & Ecker, 2020) which provide examples of figures that can be used as starting points or templates for new users (also see Section\@ref(brain-code-a-web-app-for-generating-code-templates-for-brain-visualizations)). As the popularity of code-sharing for figure increases, there will be a cornucopia of templates that can be used as the basis for new figures.

While brain visualizations are often thought of as the end results of analyses, they also form a vital part of quality control for imaging data. Tools to automatically detect artefacts, de-noise the data and generate derivatives are becoming more robust, but we are not yet at the stage where visualizing the data during processing is no longer necessary. Nonetheless, when working with large datasets such as Human Connectome Project (Van Essen et al., 2013) or UK BioBank (Sudlow et al., 2015), it is simply not feasible to use traditional GUI-based tools to visually examine the data. The time it takes to open a single file and achieve the desired visualization settings vastly compounds when working with large datasets. Knowing how to programmatically generate brain visualizations can allow for iteration of visualization code over each image of a large datasets making checking the quality of each data processing step achievable. The visual outputs of each iteration can be complied into accessible documents which can be easily scrolled, with more advanced usage allowing for the creation of interactive HTML reports (see Section \@ref(generate-and-share-visualization)), similar to those created by standardized data processing tools like *fmriprep* (Esteban et al., 2019). Increasing capacity to conduct visual quality control on larger datasets will improve the identification of processing errors and result in more reliable and valid findings.

@@

Often in neuroimaging studies, programming languages such as R, Python and MATLAB are used for statistical analysis and generating non-brain figures, but the brain figures are outsourced to separate GUI-based tools such as *FSLeyes*, *Freeview* or *ITK-snap*. Increasingly popular software such as *R Markdown*, *Quarto* and *Jupyter Notebook* allow for the mixing of prose and code in a single script, resulting in fully reproducible and publication ready papers. By using code-based tools available within the preferred environment, brain visualizations can be directly integrated and embedded within a paper or report. For instance, a fully reproducible version of the current paper can be found on [GitHub](https://github.com/sidchop/RepoNeuroVis). Some journals that publish neuroimaging studies are moving towards allowing the submission of reproducible manuscripts, including reproducible figures (e.g. *e-Life*, *Aperture Neuro*), with other journals like *F1000Research* and *GigaScience* even allowing on-demand re-running of code linked to the associated article via cloud-based platforms like *Code Ocean* (Code Ocean, 2021).

Neuroimaging data are often spatially 3D and can have multiple time points, adding a 4th dimension (e.g., functional imaging data). Thus, communicating findings or evaluating quality using static 2D slices is challenging, and may not be the best representation of the data, or the associated interpretations. While well-curated 3D renderings can help with spatial localisation (see Madan, 2015; Pernet & Madan, 2019), in the end, static images can only provide an incomplete representation of the data, and they force researchers to choose the "best" angle or slice to show, which often involves compromising one result to emphasize another. An added advantage of some of the code-based tools is the generation of 'rich' media like interactive figures or animations, which allow users to zoom, rotate and scroll through slices. Interacting with a figure in this way can improve scientific communication of findings. Linking to or even embedding these videos or interactive figures in papers can greatly enhance the communication of findings and make papers more engaging for the reader. Such rich brain visualizations lend themselves to being embedded or shared on science communication mediums beyond academic papers - such as presentations, websites and social media - all of which can promote the communication of research with peers and reach larger audiences (Li and Xie, 2020). This last point is becoming increasingly salient as promoting science on social media has become a core medium for spreading discoveries via science communication to the public (Mueller-Herbst, et al., 2020; Smith & Seitz, 2019; Huber et al., 2019), to the research community (Luc et al., 2021; Quintana, 2020a), and even a primary avenue for employment opportunities for early-career researchers (Baker, 2015; Lee, 2019). Overall, public engagement is one of the cornerstones of science, and the images we create are at the center of the process.

@@

## Selecting your programming language

The first step (Figure 1) in generating code-based visualization is selecting the coding language you want to use. Three of the most popular languages are R, Python and MATLAB, all of which have many options for generating brain visualizations (see Table 1). This decision can be made based on what language you have prior experience with, or if you used one of these languages for the analysis part of your project, there can be advantages in using the same language for visualizations. Switching from your analysis environment to separate visualization environment, or a GUI-based visualization software, can be a cumbersome deviation from the scientific workflow. This can make debugging errors more difficult, as you must regularly switch program to visually examine the results of any modifications or adjustments to prior analyses. Using the brain visualization tools that already exist within your chosen programming environment can provide instant visual feedback on the impact of modifications to processing or analysis. Alternatively, the choice of programming language may be dictated by visualization and data-type you require. For instance, visualizing streamlines from tractography may not be currently available in the R environment (Table 1), and therefore requires the use of Python or MATLAB. Other constraints may include access to proprietary software like MATLAB, which would necessitate the use of open-source options such as R or Python.

@@

## Identify your visualization type

Neuroimaging data and its derivatives can be visualized in multiple forms, which have different associated file-types and visualization requirements (see Section \@ref(identify-your-input-file-format.)). A brief description of more popular visualization types is provided below:

***Voxel*****.** A voxel is the 3D equivalent of a pixel. In neuroimaging, voxels are used to represent the intensity values of a 3D scan, such as an MRI or CT scan. The voxels can be rendered in different colors to indicate different tissue types or other features of interest. For example, in functional MRI (fMRI) scans, voxels can be colored based on their level of activation, to show which areas of the brain are more active during a specific task or for MRI, parts of the brain showing statistically significant anatomical alterations. These visualizations are often displayed as either slices in axial, sagittal or coronal planes (Fig2A-B; Fig3A-B; or a 3D rendering of the whole brain. Statistical values are displayed as overlays on template anatomical images, which follow a common stereotaxic coordinate system (e.g., MNI152), or on individual-specific anatomical images ('native-space').

***Vertex.*** A vertex is a point in 3D space that represents a location on the surface. In neuroimaging, vertices are used to create a mesh representation of a brain structure, such as the cerebral cortex or a subcortical structure. Each vertex has a set of coordinates that specify its location in 3D space and is connected to other vertices to form triangles, which make up the mesh. Vertices can be used to create a 3D visualization of the brain surface and color-coded based on different attributes such as sulcal depth, thickness, or functional activation. These visualizations are often displayed as 3D rendering of each hemisphere from medial and lateral views (Fig2C,G-H; Fig3C-D). Statistical values are displayed as overlays on template surfaces which follow a common stereotaxic coordinate system and fixed number of vertices (e.g., average; Fig2C,G; Fig3C-D), or on individual-specific surfaces that have been reconstructed using an anatomical image (Fig2H).

***Regions of Interest (ROI).*** In neuroimaging, ROIs are used to identify specific brain structures or areas that are relevant to the research question. ROIs can be defined using various methods, such as manual tracing, atlas-based parcellation, or functional activation patterns. Visualizing ROIs can be done by grouping and assigning the same statistical value or color to sets of voxels or vertices (Fig2B-C; Fig3B-C), or can be done through polygons. Polygonal brain visualizations are simple 2D or 3D shapes which graphically represent the brain or specific structures, but do not carry any additional information on spatial coordinates and only roughly estimate the shape of the brain and its structures. Each region of 2D (Fig 2D,F) and 3D (Fig 2E; Fig3E) polygon visualizations can be filled in with colors representing a region label or a statistical value.

***Edge.*** In the context of neuroimaging, an edge often represents a connection between two brain regions, which can be thought of as a line connecting two nodes (brain regions) in a network. To visualize edges, one common method is to use a matrix representation, often called a 'connectivity' or 'adjacency' matrix, which shows the strength of the connections between all pairs of brain regions. These matrices can be displayed as a heat map, where different colors represent different levels of connectivity. Another way to represent edges is using a 2D graph, where the vertices represent brain regions and the edges represent the connections between those regions, which displayed in a variety of ways, such as a 2D or 3D representation of the brain (Fig2I; Fig3F), with edges represented as lines connecting the vertices. The edges and nodes can also be color-coded based on properties such as the strength of the connection.

***Streamlines.*** Streamlines are a representation of the white matter fibers in the brain and are usually generated using tractography applied to diffusion MRI (dMRI). Streamlines are typically visualized as 3D lines that connect different points of the brain and can be used to create 3D visualizations of specific white matter tracts, or all streamlines between regions. They can be color-coded based on the direction of the fibers or properties of the tract, such as myelination. These visualizations can be overlaid on voxel or vertex level anatomical brain representations to provide an anatomical reference point.

The following two figures provide examples of voxel, vertex, ROI and edge-level visualizations generated within R (Figure 2) and Python (Figure 3) using open source, well documented and beginner-friendly packages. These are not an exhaustive representation of packages available for visualizing brain data in R and Python (see Table 1). Rather, the figures aim to give the reader a sense of the many options available, and an entry point to choosing the type of brain visualization needed (also see Section 4). All code used to compile the figures, as well as the contents of each panel are provided in the accompanying [online repository](https://github.com/sidchop/RepoNeuroVis).

@@

```{=tex}
\newpage
\blandscape
```
```{r Figure3, echo=FALSE, out.height = '70%', out.width = '90%', fig.align='center', dev='cairo_pdf'}
knitr::include_graphics("./data/python/Figure2.pdf")
```

@@

## Identify your input file-format

Human neuroimaging data and its derivatives come in many different file formats, and code-based visualization packages may have specific requirements about which formats they are designed to work with. Table 1 provides a key, listing which data-type can be used as inputs for each package. Some of the most common MRI file-formats used for visualizations are briefly described below:

***Plain text format.*** The simplest input formats are a scalar, vector, and matrix, which represent a single data-point, one-dimensional array of data (e.g., a single column or row), and a two-dimensional array of data (e.g. multiple columns or rows), respectively. These data are often stored in plain text formats such as `.txt`, `.csv`, and `.tsv` and are often generated by neuroimaging analyses software such as SPM, FSL or FreeSurfer. These files can also contain rows and columns of region names and/or spatial coordinates. All programming languages have functions to read these plain text formats into your environment as data structures such as matrices and arrays. If you have conducted your imaging analyses using your chosen programming language, rather than stand-alone software, the data you want to visualize will likely already be available to you within your coding environment. These formats are often used in region-level visualizations, where groups of voxels or vertices share the same value or color (Fig2B-F; Fig3B-E), or in edge-level visualizations, where a matrix is used to identify which two brain regions are connected by an edge (Fig2F; Fig3F).

***NIfTI*****.** NIfTI (`.nii`) files store 3D or 4D image data, which is often a matrix of voxel intensities or a series of 3D matrices for 4D data such as fMRI and dMRI. The image data is stored as a 3D matrix of voxel intensities, and the header contains information such as the image dimensions, voxel size, and data type. Additional information such as patient demographics and scanner parameters can also be stored in the header in the form of metadata.

***GIfTI.*** GIFTI (`.gii`) is an extension of the NifTI format, which stores data in a surface-based format, with the data represented as a set of vertices, edges, and faces that define a surface mesh. The format also includes support for storing data such as curvature, thickness, and functional activity maps on the surface mesh. Often `.gii` files will have a pre-indicator of what information the file contains, such as `.surf.gii`, which would contain only vertices, edges and faces to define a surface mesh. Or `.func.gii` files which contain data values for every vertex, which are essentially data arrays whose indices correspond to a surface file and need a matching surface file to know where in the brain to assign the data values. GIFTI files can also store multiple surfaces in a single file, and can include information about the topology of the surfaces, such as the number of vertices, edges, and faces. Additional metadata can also be stored in the header.

***CIfTI.*** CIFTI (`.cii`) files can store data from both surface-based and volume-based neuroimaging analyses. For surface-based data, the file contains vertex coordinates, and data values at each vertex. For volume-based data, the file contains a 3D matrix of voxel intensities. Often this datatype is used to represent the cortex as vertices, and subcortical structures as voxels. CIFTI files can be divided into three main types: `.dtseries` (store time-series data, such as fMRI data), `.dtscalar` (store scalar data, such as thickness or curvature maps), and `.dtlabel` (store label data, such as parcellations of the brain). CIFTI files can also include fields with additional information such as surface and volume registration information.

>>>>

---
title: A Practical Guide for Generating Reproducible and Programmatic Neuroimaging Visualizations
authors:
  - name: Sidhant Chopra
    thanks: Corresponding author
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
    email: sidhant.chopra@yale.com
  - name: Loïc Labache
    department: Department of Psychology
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Elvisha Dhamala
    department: Institute of Behavioral Sciences
    affiliation: Feinstein Institutes for Medical Research,
    location: NY, USA 11030
  - name: Edwina R Orchard
    department: Yale Child Study Center
    affiliation: Yale University
    location: CT, USA, 06511
  - name: Avram Holmes
    department: Department of Psychiatry, Brain Health Institute
    affiliation: Rutgers University
    location: NJ, USA, 08854
keywords:
  - Neuroimaging visualization
  - Reproducibility 
  - Programmatic figures
  - Open Science
  - R
  - Python
  - Brain Visualization
abstract: Neuroimaging visualizations form the centerpiece of the interpretation and communication of scientific results, and are also a cornerstone for data quality control. Often, these images and figures are produced by manually changing settings on Graphical User Interfaces (GUIs). There now exist many well-documented code-based brain visualization tools that allow users to programmatically generate publication-ready figures directly within R, Python and MATLAB environments. Here, we provide a rationale for the wide-spread adoption of code-generated brain visualizations by highlighting corresponding advantages in replicability, flexibility, and integration over GUI based tools. We then provide a practical guide outlining the steps required to generate these code-based brain visualizations. We also present a comprehensive table of tools currently available for programmatic brain visualizations (https://sidchop.shinyapps.io/braincode_selector/) and provide didactic examples of visualizations and associated code as a point of reference. Finally, we provide a web-app (https://sidchop.shinyapps.io/braincode/) that can generate simple code-templates as starting points for these visualizations.
output: 
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
    keep_tex: true
    latex_engine: lualatex
  bookdown::word_document2:
    toc: false
linestretch: 1.5
geometry: "left=1.5cm, right=1.5cm, top=1.5cm, bottom=1.5cm"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{setspace}\onehalfspacing
- \usepackage[left]{lineno}
- \usepackage[normalem]{ulem} 
- \modulolinenumbers[5]
- \linenumbers
---

@@

The visualization of neuroimaging data is one of the primary ways in which we evaluate data quality, interpret results, and communicate findings. These visualizations are commonly produced using graphical user interface-based (GUI) tools where individual images are opened and, within each instance, display settings are manually changed until the desired output is reached. In large part, the choice to use GUI-based software has been driven by a perception of convenience, flexibility, and accessibility. However, there now exist code-based software packages that are well-documented and do not require high-level and comprehensive knowledge of programming, making them accessible to the neuroimaging community. These tools are flexible and allow for the generation of reproducible, high-quality, and publication-ready brain visualizations in only a few lines of code, especially within the R, Python and MATLAB environments. Here, we first discuss the rationale for the widespread adoption of code-generated visualizations by highlighting major advantages in replicability, flexibility, and integration over GUI based tools. We then provide a practical guide outlining the steps required to make code-based brain visualizations and provide a web-app (<https://sidchop.shinyapps.io/braincode/>; see Section\@ref(brain-code-a-web-app-for-generating-code-templates-for-brain-visualizations)) that can generate simple code-templates as starting points for these visualizations. We also present a comprehensive table of tools currently available for programmatic brain visualizations (Table 1) and provide instructive examples of visualizations and associated code as a point of reference (Figure 2-3). The focus of this guide is on human brain magnetic resonance imaging (MRI) data, but many of the principles discussed and tools provided will equally apply to visualizing data from other organs and imaging modalities such as EEG, MEG, PET and CT.

@@

Notably, some GUI-based tools have historically offered command-line access to generate replicable visualizations (e.g., FreeView, FSLeyes, surfice), making their use potentially equally replicable to purely code-based tools. Use of these specialized command-line interfaces can provide a useful middle ground for those who have little experience with coding environments. Nonetheless, these interfaces often still have a learning curve, but can lack other advantages, such as iteration, provided by the preferred programming environment (see Sections \@ref(flexibility-and-scalability) and \@ref(integrative-and-interactive-reporting)). Likewise, other GUI-based tools offer replicability in the form of automatically generated batch scripts (text files containing lines of specialized commands that can be re-executed) or in-built terminals, which can be idiosyncratic and may lack documentation to make them easily usable or replicable by those not familiar with the specific software.

## Flexibility and scalability

Being able to exactly replicate a figures via code has marked advantages beyond open science practices. In particular, the ability to reprogram inputs (such as statistical maps) and settings (such as color schemes, thresholds, and visual orientations) can streamline the entire scientific workflow. Changing inputs and settings via code allows for the easy production of multiple figures, such as those resulting from multiple analyses that require similar visualizations. A simple for-loop or plotting function with altered input and/or settings-of-interest can be a powerful method for exploring visualization options or rapidly creating multi-panel figures. Likewise, an arduous request from a reviewer or collaborator to alter the image processing or analysis becomes less of a burden when the associated figures can be re-generated with a few lines of code, as opposed to re-pasting and re-illustrating them manually. Having a code-base with modifiable inputs can mean that the generation of visualizations requires less time, energy and effort than image and instance specific GUI-based generation. This also makes it easier to generate consistent figures across subsequent projects. Keep in mind that the gains of writing code for figures are cumulative, and in addition to improving programming skills, one can build a code-base for figure generation that can continue to be reused and shared throughout a scientific career.

Precise controls via code over visualization settings, such as color schemes, legend placement and camera angles, can provide much greater flexibility over visualizations. Nonetheless, part of the appeal of GUI-based tools is that the presets for such settings can provide a useful starting point and reduce the decision burden on novice users. However, similar presets are often available in the form of default settings across most code-based packages, negating the need for the user to manually enter each and every choice required for creating an image. Most code-based tools also come with documentation, with R-packages on the *CRAN* or *Neuroconductor* (Muschelli et al., 2019) repositories strictly requiring detailed guidance. Recent packages have started to include detailed beginner-friendly documentation in GitHub repositories, or even entire papers (e.g., Pham, Muschelli, & Mejia, 2022; Mowinckel & Vidal-Piñeiro, 2020; Huntenburg et at., 2017; Schäfer & Ecker, 2020) that provide examples of figures that can be used as starting points or templates for new users (also see Section\@ref(brain-code-a-web-app-for-generating-code-templates-for-brain-visualizations)). As the popularity of code-sharing for figure increases, there will be a cornucopia of templates that can be used as the basis for new figures.

While brain visualizations are often thought of as the end results of analyses, they also form a vital part of quality control for imaging data. Tools to automatically detect artefacts, de-noise the data and generate derivatives are becoming more robust, but we are not yet at the stage where visualizing the data during processing is no longer necessary. Nonetheless, when working with large datasets such as Human Connectome Project (Van Essen et al., 2013) or UK BioBank (Sudlow et al., 2015), it is simply not feasible to use traditional GUI-based tools to visually examine the data. The time it takes to open a single file and achieve the desired visualization settings vastly compounds when working with large datasets. Knowing how to programmatically generate brain visualizations can allow for iteration of visualization code over each image of a large datasets making quality checks of each data processing step achievable. The visual outputs of each iteration can be complied into accessible documents which can be easily scrolled, with more advanced usage allowing for the creation of interactive HTML reports (see Section \@ref(generate-visualization)), similar to those created by standardized data processing tools like *fmriprep* (Esteban et al., 2019). Increasing capacity to conduct visual quality control on larger datasets will improve the identification of processing errors and result in more reliable and valid findings.

@@

Often in neuroimaging studies, programming languages such as R, Python and MATLAB are used for statistical analysis and generating non-brain figures, but the brain figures are outsourced to separate GUI-based tools such as *FSLeyes*, *Freeview* or *ITK-snap*. Increasingly popular software such as *R Markdown*, *Quarto* and *Jupyter Notebook* allow for the mixing of prose and code in a single script, resulting in fully reproducible and publication ready papers. By using code-based tools available within the preferred environment, brain visualizations can be directly integrated and embedded within a paper or report. For instance, a fully reproducible version of the current paper can be found on [GitHub](https://github.com/sidchop/RepoNeuroVis). Some journals that publish neuroimaging studies are moving towards allowing the submission of reproducible manuscripts, including reproducible figures (e.g. *eLife*, *Aperture Neuro*), with other journals like *F1000Research* and *GigaScience* even allowing on-demand re-running of code linked to the associated article via cloud-based platforms like *Code Ocean* (Code Ocean, 2021).

Neuroimaging data are often spatially 3D and can have multiple time points, adding a 4th dimension (e.g., functional imaging data). Thus, communicating findings or evaluating quality using static 2D slices is challenging, and may not be the best representation of the data, or the associated interpretations. While well-curated 3D renderings can help with spatial localisation (see Madan, 2015; Pernet & Madan, 2019), in the end, static images can only provide an incomplete representation of the data, and they force researchers to choose the "best" angle or slice to show, which often involves compromising one result to emphasize another. An added advantage of some of the code-based tools is the generation of 'rich' media like interactive figures or animations, which allow users to zoom, rotate and scroll through slices. Interacting with a figure in this way can improve scientific communication of findings. Linking to or even embedding these videos or interactive figures in papers can greatly enhance the communication of findings and make papers more engaging for the reader. Such rich brain visualizations lend themselves to being embedded or shared on science communication mediums beyond academic papers---such as presentations, websites and social media---all of which can promote the communication of research with peers and reach larger audiences (Li and Xie, 2020). This last point is becoming increasingly salient as promoting science on social media has become a core medium for spreading discoveries via science communication to the public (Mueller-Herbst, et al., 2020; Smith & Seitz, 2019; Huber et al., 2019), to the research community (Luc et al., 2021; Quintana, 2020a), and even a primary avenue for employment opportunities for early-career researchers (Baker, 2015; Lee, 2019). Overall, public engagement is one of the cornerstones of science, and the images we create are at the center of the process.

@@

## Selecting a programming language

The first step (Figure 1) in generating code-based visualization can be selecting the coding language. Three of the most popular languages are R, Python and MATLAB, all of which have many options for generating brain visualizations (see Table 1). This decision can be made based on what language the user has prior experience with, or if one of these languages was for the analysis part of a project. There can be advantages in using the same language for visualizations and analyses, as switching from analysis environment to separate visualization environment, or a GUI-based visualization software, can be a cumbersome deviation from the scientific workflow. This can make debugging errors more difficult, as the user must regularly switch program to visually examine the results of any modifications or adjustments to prior analyses. Using the brain visualization tools that already exist within a chosen programming environment can provide instant visual feedback on the impact of modifications to processing or analysis. Alternatively, the choice of programming language may be dictated by visualization and data-type required. For instance, visualizing streamlines from tractography may not be currently available in the R environment (Table 1), and therefore requires the use of Python or MATLAB. Other constraints may include limited access to proprietary software like MATLAB, which would necessitate the use of open-source options such as R, Python or Octave.

@@

## Identify a visualization type

Neuroimaging data and its derivatives can be visualized in multiple forms, that have different associated file-types and visualization requirements (see Section \@ref(identify-input-file-format)). A brief description of more popular visualization types is provided below:

***Voxel*****.** In neuroimaging, voxels are used to represent the intensity values of a 3D scan, such as an MRI or CT scan. The voxels can be rendered in different colors to indicate different tissue types or other features of interest. For example, in functional MRI scans, voxels can be colored based on their level of activation, to show which areas of the brain are more active during a specific task. These visualizations are often displayed as either slices in axial, sagittal or coronal planes (Fig2A-B; Fig3A-B; or a 3D rendering of the whole brain. Statistical values are displayed as overlays on template anatomical images, which follow a common stereotaxic coordinate system (e.g., MNI152), or on individual-specific anatomical images.

***Vertex.*** In neuroimaging, vertices are used to create a mesh representation of a brain structure, such as the cerebral cortex or a subcortical regions. Each vertex has a set of coordinates that specify its location in 3D space and is connected to other vertices to form triangles, which make up the mesh. Vertices can be used to create a 3D visualization of the brain surface and color-coded based on different attributes such as sulcal depth, thickness, or functional activation. These visualizations are often displayed as 3D rendering of each hemisphere from medial and lateral views (Fig2C,G-H; Fig3C-D). Statistical values are displayed as overlays on template surfaces which follow a common stereotaxic coordinate system and fixed number of vertices (e.g., fsaverage; Fig2C,G; Fig3C-D), or on individual-specific surfaces that have been reconstructed using an anatomical image (Fig2H).

***Regions of Interest (ROI).*** In neuroimaging, ROIs are used to identify specific brain structures or areas that are relevant to the research question. ROIs can be defined using various methods, such as manual tracing, atlas-based parcellation, or functional activation patterns. Visualizing ROIs can be done by grouping and assigning the same statistical value or color to sets of voxels or vertices (Fig2B-C; Fig3B-C), or can be done through polygons. Polygonal brain visualizations are simple 2D or 3D shapes which graphically represent the brain or specific structures, but do not carry any additional information on spatial coordinates and only roughly estimate the shape of the brain and its structures. Each region of 2D (Fig 2D,F) and 3D (Fig 2E; Fig3E) polygon visualizations is filled in with colors indicating a region label or a statistical value.

***Edge.*** In the context of neuroimaging, an edge often represents a connection between two brain regions, which can be visualised as a straight or curved line connecting two nodes (brain regions) in a network. One common way data is organized for edge-level visualization is a matrix, often called a 'connectivity' or 'adjacency' matrix, which indexes the presence and strength of the connections between pairs of brain regions. Visualitaion tools often convert these matrices into network graphs, where the vertices represent brain regions and the edges represent the connections between those regions, and can be displayed in a variety of ways, such as overlayed on a 2D or 3D representation of the brain (Fig2I; Fig3F), with edges represented as straight lines connecting the vertices. Often, visual properties of edges and nodes can also be adjusted to convey information, such as sized or color-coded based the strength or number of connections.

***Streamlines.*** While similar to edges, streamlines are specifically used to representation of the white matter fibers in the brain and are usually an out put of applying tractography applied to diffusion-weighted MRI. Streamlines are typically visualized as curved 3D lines that connect different points of the brain and can be used to create 3D visualizations of specific white matter tracts, or all streamlines between regions. They can be color-coded based on the direction of the fibers or properties of the tract, such as myelination. These visualizations can be overlaid on voxel or vertex level anatomical brain representations to provide an anatomical reference point.

The following two figures provide examples of voxel, vertex, ROI and edge-level visualizations generated within R (Figure 2) and Python (Figure 3) using open source, well documented and beginner-friendly packages. These are not an exhaustive representation of packages available for visualizing brain data in R and Python (see Table 1). Rather, the figures aim to give the reader a sense of the many options available, and an entry point to choosing the type of brain visualization needed (also see Section\@ref(brain-code-a-web-app-for-generating-code-templates-for-brain-visualizations)). All code used to compile the figures, as well as the contents of each panel are provided in the accompanying [online repository](https://github.com/sidchop/RepoNeuroVis).

@@

```{=tex}
\newpage
\blandscape
```
```{r Figure3, echo=FALSE, out.height = '60%', out.width = '85%', fig.align='center', dev='cairo_pdf'}
knitr::include_graphics("./data/python/Figure2.pdf")
```

@@

## Identify input file-formats

Neuroimaging data and its derivatives come in many different file formats, and code-based visualization packages have specific formats they are designed to work with. Table 1 provides a key, listing which data-type can be used as inputs for each package. Some of the most common MRI file-formats used as inputs for visualizations are briefly described below:

***Plain text format.*** The simplest input formats are a scalar, vector, and matrix, which represent a single data-point, one-dimensional array of data (e.g., a single column or row), and a two-dimensional array of data (e.g. multiple columns or rows), respectively. These data are often stored in plain text formats such as `.txt`, `.csv`, and `.tsv` and can be generated by neuroimaging analyses software such as SPM, FSL or FreeSurfer. These files can also contain rows and columns of region names and/or spatial coordinates. All programming languages have functions to read these plain text formats into the coding environment. These formats are often used in region-level visualizations, where groups of voxels or vertices share the same value or color (Fig2B-F; Fig3B-E), or in edge-level visualizations, where a matrix is used to identify regions are connected by an edge (Fig2F; Fig3F).

***NIfTI*****.** NIfTI (`.nii`) files store 3D or 4D image data, which is often a matrix of voxel intensities or a series of 3D matrices for 4D data such as fMRI and dMRI. The image data is stored as a 3D matrix of voxel intensities, and the header contains additional information, such as the image dimensions and voxel size. Additional information such as the subjects demographics and scanner parameters can also be stored in the header in the form of metadata.

***GIfTI.*** GIFTI (`.gii`) is an extension of the NIFTI format, which stores data in a surface-based format, where data is represented as a set of vertices, edges, and faces that define a surface mesh. The format also includes support for storing data such as curvature, thickness, and functional activity maps on the surface mesh. Often `.gii` file names will have a pre-indicator of what information the file contains, such as `.surf.gii`, which would contain only vertices, edges and faces to define a surface mesh. Or `.func.gii` files which contain data values for every vertex, which are essentially data arrays whose indices correspond to a surface file and need a matching surface file to know where in the brain to assign the data values. GIFTI files can also store multiple surfaces in a single file, and can include information about the topology of the surfaces, such as the number of vertices, edges, and faces. Additional metadata can also be stored in the header.

***CIfTI.*** CIFTI (`.cii`) files can store data from both surface-based and volume-based neuroimaging analyses, combining aspects of NIFTI and GIFTI files. For surface-based data, the file contains vertex coordinates, and data values at each vertex. For volume-based data, the file contains a 3D matrix of voxel intensities. Often this datatype is used to represent the cortex as vertices, and subcortical structures as voxels. CIFTI files can be divided into three main types: `.dtseries` (store time-series data, such as fMRI data), `.dtscalar` (store scalar data, such as thickness or curvature maps), and `.dtlabel` (store label data, such as parcellations of the brain). CIFTI files can also include fields with additional metadata such as surface and volume registration information.
